---
title: "United States Energy Generation 2001-2022"
author: "Kevin Sheard, John Beall, Andrew Vo"
date: "`r Sys.Date()`"
output: slidy_presentation
runtime: shiny
source: https://www.kaggle.com/datasets/kevinmorgado/us-energy-generation-2001-2022
---
## In this data set we're looking at the 17 sustainable development goals of the UN. Within these 17 goals we saw that the UN's energy goal could be a potentially high value benefactor of data science to help curb the generation of fossil fuels and promote green energy. We chose this topic with the end goal of finding interesting discoveries that could provide actionable data to the every day consumer and help educate their decision to dictate their personal energy needs within current trends and available sources today. We hope that through our visualization of trending energy generation, production, and interactive comparisons an individual may be better informed to make their steps towards a greener future!


```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = F,
	message = FALSE,
	warning = FALSE
)
```

```{r}
# Load tidyverse for manipulating data
#Load ggplot2 for graphing data
#Load shiny and plotly for interactive data
library(tidyverse, warn.conflicts = F)
library(ggplot2, warn.conflicts = F)
library(plotly, warn.conflicts = F)
library(shiny, warn.conflicts = F)
library(ggthemes, warn.conflicts = F)
```

```{r}
#Read in the energy data
energy = read_csv("archive/organised_Gen.csv")

```

```{r}
#renaming the columns and removing the "ID" column
colnames(energy) = c("ID", "year", "month", "state", "producer", "source", "generation")
energy = energy %>% select(-`ID`)
```

## Glimpse of the Energy Dataset

```{r}
#adjusting the dates so that they can used with a time series graph
energy$Date<-as.Date(with(energy,paste(year,month,Day = 1,sep="-")),"%Y-%m-%d")
glimpse(energy)
```
- library(tidyverse) Why else? Makes things tidy
- library(ggplot2) Because graphs
- library(plotly) Because graphs with shiny
- library(shiny) Whats better than data? Interactive data!
- library(ggthemes) Gotta be fancy


## US Power Generation

```{r}
#plot a graph for US avg. power generation per year from years 2001-may 2022
energy %>% 
  filter(producer == "Total Electric Power Industry", state == "US-TOTAL") %>% 
  ggplot() + 
  geom_point(aes(x = year, y = generation/1000000, color = source)) +
  #we will generation/1000000 to create Tera Watt Hours instead Mega Watt Hours for readability for the rest of the project
  geom_smooth(aes(x = year, y = generation/1000000, color = source)) +
  labs(title = "US average power generation per year", y = "Generation(TWH)", x = "Year") + theme_clean()
  
```

## US Power Generation

```{r fig.height=8, fig.width=5}
#setting figure boundaries to ensure readability
#plot a graph for accumulative US power generation per year from years 2001-may 2022 for each state
total_energy = energy %>% filter(producer == "Total Electric Power Industry", state != "US-TOTAL") 
fig = energy %>% 
  filter(producer == "Total Electric Power Industry", state != "US-TOTAL") %>% 
  group_by(year, month, state) %>% 
  summarize(total = sum(generation)/2) %>% 
  right_join(total_energy) %>%
  filter(source != "Total") %>% 
  arrange(total) %>% 
  ggplot() +
  geom_col(aes(x = state, y = generation/1000000, fill = source)) +
  #used columns and flipped coords to create clarity on graph
  labs(title = "US Power Generation by State", y = "Generation(TWH)", x = "State") +
  coord_flip() + theme_clean()

fig
```

## Texas Power Generation

```{r}
#Display the Texas avg power generation per year from years 2001-may 2022 
energy %>% 
  filter(producer == "Total Electric Power Industry", state == "TX") %>% 
  ggplot() + 
  geom_point(aes(x = year, y = generation/1000000, color = source)) +
  geom_smooth(aes(x = year, y = generation/1000000, color = source)) +
  labs(title = "Texas average power generation per year", x = "Year", y = "Power Generation(TWH)") + theme_clean()

#Display the Texas avg power generation per month to show variation of the power usage over the course of a year
energy %>% 
  filter(producer == "Total Electric Power Industry", state == "TX") %>% 
  ggplot() + 
  geom_point(aes(x = month, y = generation/1000000, color = source)) +
  geom_smooth(aes(x = month, y = generation/1000000, color = source)) +
  labs(title = "Texas average power generation per month", x = "Month", y = "Power Generation(TWH)") + theme_clean()
```

## Texas Power Generation

```{r fig.height=20}
#choosing display height for better readability

ui <- fluidPage(
  # App title ----
  titlePanel("Texas Total Power Generation Over Time"),
  # Sidebar layout with a input and output definitions ----
  sidebarLayout(
    # Sidebar panel for inputs ----
    sidebarPanel(
      position = "right",
      # checkbox for source
      checkboxGroupInput("sources", 
        label = h3("Source:"), 
        choices = unique(energy$source),
        selected = "Total"),
      # checkbox to determine facet status
      checkboxInput("facetStatus",
                    "Facet",
                    value = TRUE),
      # input selection for what state to graph on the top graph
      selectInput("state1",
                  label = "State(top)",
                  choices = unique(energy$state)),
      # input selection for what state to graph on the bottom graph
      selectInput("state2",
                  label = "State(bottom)",
                  choices = unique(energy$state))
    ),

    # Main panel for displaying outputs ----
    mainPanel(
      #output the top plot
      plotlyOutput("totalGenPlot",
                   width = "100%",
                   height = "100%"),
      #output the bottom plot
      plotlyOutput("totalGenPlot2")
    )
  )
)

server = function(input, output){
  #Use reactive so that we can interact with the shiny
  shinyEnergy = reactive({
    energy %>% filter(producer == "Total Electric Power Industry", state == input$state1, year != 2022, source == input$sources) %>% group_by(Date, source) %>% 
    summarize(total_generation = sum(generation)) %>% mutate(total_generation = total_generation / 1000000)
  })
  
  shinyEnergy2 = reactive({
    energy %>% filter(producer == "Total Electric Power Industry", state == input$state2, year != 2022, source == input$sources) %>% group_by(Date, source) %>% 
    summarize(total_generation = sum(generation)) %>% mutate(total_generation = total_generation / 1000000)
  })

  output$totalGenPlot = renderPlotly({
    #using plotly so Shiny behaves with our graph, and using fig2 as an object so that we can call up a way to use the Shiny
    fig2 = shinyEnergy() %>% 
    ggplot() +
    geom_line(aes(x= Date, y= total_generation, color = source)) +
    geom_smooth(aes(x = Date, y= total_generation, color = source)) +
    labs(x = "Date", y = "Power Generation(TWH)") 
    # facetEnergy is reactive so we can change its status, has source input and changes title based on the state check box
    facetEnergy = reactive({
      if(input$facetStatus == T){
        fig2 = fig2 + facet_wrap(~source)
      }
      else{
        fig2
      }
      fig2 = fig2 + labs(title = paste(input$state1, "Total Power Generation Over Time"))
    })
    
    fig2 = facetEnergy()
    #create the plotly
    ggplotly(fig2, tooltip = c("x", "y", "source"))
  })
  #second graph is constructed similarly to the graph above
  output$totalGenPlot2 = renderPlotly({
    fig2 = shinyEnergy2() %>% 
    ggplot() +
    geom_line(aes(x= Date, y= total_generation, color = source)) +
    geom_smooth(aes(x = Date, y= total_generation, color = source)) +
    labs(x = "Date", y = "Power Generation(TWH)") 
    
    facetEnergy = reactive({
      if(input$facetStatus == T){
        fig2 = fig2 + facet_wrap(~source)
      }
      else{
        fig2
      }
      fig2 = fig2 + labs(title = paste(input$state2, "Total Power Generation Over Time"))
    })
    
    fig2 = facetEnergy()
    ggplotly(fig2, tooltip = c("x", "y", "source"))
  })
}

shinyApp(ui = ui, server = server)
```

## Interesting Outliers

```{r}
#Noticed interesting outliers of data, and displayed it to show the magnitude of negative values of power generation.
energy %>% 
  filter(producer == "Total Electric Power Industry", state != "US-TOTAL", generation < -5000) %>% 
  ggplot() +
  geom_col(aes(x = state, y = generation/1000000, fill = source)) +
  labs(title = "States with negative power generation (<-5000)", x ="State", y="Power Generation(TWH)") + theme_clean()
```

## US Power Generation

```{r}
#second shiny to display power generation for the US 
ui <- fluidPage(
  # App title ----
  titlePanel("US Total Power Generation Over Time"),
  # Sidebar layout with a input and output definitions ----
  sidebarLayout(
    # Sidebar panel for inputs ----
    sidebarPanel(
      position = "right",
      # checkbox for different sources
      checkboxGroupInput("sources", 
        label = h3("Source:"), 
        choices = unique(energy$source),
        selected = "Total"),
      # checkbox for whether to facet graph or not
      checkboxInput("facetStatus",
                    "Facet",
                    value = TRUE),
  
      hr(),
      fluidRow(column(3, verbatimTextOutput("value")))
    ),

    # Main panel for displaying outputs ----
    mainPanel(
      # Output: Verbatim text for data summary ----
      verbatimTextOutput("summary"),
      plotlyOutput("totalGenPlot")
    )
  )
)

server = function(input, output){
  # reactive in order to get what source to filter by
  shinyEnergy = reactive({
    energy %>% filter(producer == "Total Electric Power Industry", state != "US-TOTAL", year != 2022, source == input$sources) %>% group_by(Date, source) %>% 
    summarize(total_generation = sum(generation)) %>% mutate(total_generation = total_generation / 1000000)
  })
  
  
  #render the plot
  output$totalGenPlot = renderPlotly({
    fig2 = shinyEnergy() %>% 
    ggplot() +
    geom_line(aes(x= Date, y= total_generation, color = source)) +
    geom_smooth(aes(x = Date, y= total_generation, color = source)) +
    labs(title = "Total Power Generation Over Time", x = "Date", y = "Power Generation(TWH)") 
    
    # reactive function to get facet status
    facetEnergy = reactive({
      if(input$facetStatus == T){
        fig2 + facet_wrap(~source)
      }
      else{
        fig2
      }
    })
    
    fig2 = facetEnergy()
    # create a plotly from ggplot
    ggplotly(fig2, tooltip = c("x", "y", "source"))
  })
}

shinyApp(ui = ui, server = server)
```

## Producer Data Analysis

```{r}
#tables to display producer columns and energy sources
energy %>% count(producer)

energy %>% count(source)
```

## Texas/US Power Distribution Producer Type
```{r}
# US Power Source Distribution Per Producer Type. 
## This is meant to show the level of usage among energy source types used within US Energy industry categories. 
energy %>%
  filter(producer != "Total Electric Power Industry", source!="Total") %>% 
  ggplot() + 
  geom_col(aes(x = producer, y = generation/1000000, fill = source)) +
  labs(title = "US Power Distribution", x= "Producer", y= "Power Generation(TWH)") +
  theme(axis.text.x = element_text(angle = -25, vjust = .5, hjust = .1)) 

# TX Power Source Distribution Per Producer Type.
## This is meant to display the same comparison as the US graph, but in Texas. 
energy %>% 
  filter(producer != "Total Electric Power Industry", state == 'TX', source != 'Total') %>%
  ggplot() + 
  geom_col(aes(x = producer, y = generation/1000000, fill = source)) +
  labs(title = "Texas Power Distribution", x= "Producer", y= "Power Generation(TWH)") +
  theme(axis.text.x = element_text(angle = -25, vjust = .5, hjust = .1)) 
```

## Source Power Generation Per Year
```{r}
# US Power Source Generation Per Year
## This is meant to show the change of power sources used in the US on a chronological scale. 
energy %>% 
  filter(producer != "Total Electric Power Industry", state == "US-TOTAL", source != 'Total', year <= 2021) %>% 
  ggplot() + 
  geom_col(aes(x = year, y = generation/1000000, fill = source)) +
  labs(title = "US Power Generation Per Year", x = "Year", y = "Power Generation(TWH)")

# TX Power Source Generation Per Year
## This is meant to show the change of power sources used on a chronological scale in Texas.
energy %>% 
  filter(producer != "Total Electric Power Industry", state == "TX", source != 'Total', year <= 2021) %>% 
  ggplot() + 
  geom_col(aes(x = year, y = generation/1000000, fill = source)) +
  labs(title = "TX Power Generation Per Year", x = "Year", y = "Power Generation(TWH)")
```


## Total Power Distribution Over Months
```{r}
# Total US power distribution over months
## Graph is meant to create a visual on power generation over months and used as a control for comparison for states and their power generation over months. This is to set precedent for the next pair of graphs. 
energy %>% 
  filter(producer == "Total Electric Power Industry", state == 'US-TOTAL', source == 'Total') %>% 
  ggplot(aes(group = month, x = month, y = generation/1000000, fill = month))+
  geom_boxplot() +
  scale_x_continuous(breaks = c(1:12)) + labs(title= "Total US Power distribution over months", x= "Months", y="Power Generation(TWH)" ) 


# TX power distribution over months
## Graph is meant to show Texas power usage in comparison to US. This is to set precedent for the next pair of graphs.
energy %>% 
  filter(producer == "Total Electric Power Industry", state == 'TX', source == 'Total') %>% 
  ggplot(aes(group = month, x = month, y = generation/1000000, fill = month))+
  geom_boxplot() +
  scale_x_continuous(breaks = c(1:12))+
  labs(title = 'TX Power Distribution Over Months', x= "Months", y="Power Generation(TWH)") 
```

## Power Distributions over months for 2021
```{r}
# US Power Distributions over months for 2021
## This graph shows this use of power sources in the US over a twelve month period in the year 2021.
energy %>% 
  filter(producer != "Total Electric Power Industry", state == 'US-TOTAL', source != 'Total', year == 2021) %>% 
  ggplot()+
  geom_col(aes(x = month, y = generation/1000000, fill = source)) +
  labs(title = 'US Power Distribution Per Month', x= "Months", y="Power Generation(TWH)")+
  scale_x_continuous(breaks = c(1:12))
# TX Power Distributions over months for 2021
## This graph shows this use of power sources in the TX over a twelve month period in the year 2021.
energy %>% 
  filter(producer != "Total Electric Power Industry", state == 'TX', source != 'Total', year == 2021) %>% 
  ggplot()+
  geom_col(aes(x = month, y = generation/1000000, fill = source)) +
  labs(title = 'Texas Power Distribution Per Month', x= "Months", y="Power Generation(TWH)")+
  scale_x_continuous(breaks = c(1:12))
```

## Energy Sources amoung producer types US

```{r}
# US Energy Sources Among Producer Types in 2010
## This graph is meant to be used in comparison to the graph proceeding it 
## displaying the power source distributions among energy producers in the US in 2010
energy %>% 
  group_by(source, producer) %>% 
  filter(source != 'Total', year == 2010) %>% 
  ggplot() + 
  geom_col(aes(x = producer, y = generation/1000000, fill = source)) +
  labs(title = "US Energy Sources Among Producer Types in 2010", x= "Producer", y="Power Generation(TWH)")+
  theme(axis.text.x = element_text(angle = -15, vjust = 1, hjust = .1)) 

# US Energy Sources Among Producer Types in 2020
## This graph is meant to be used in comparison to the graph proceeding it 
## displaying the power source distributions among energy producers in the US in 2020
energy %>% 
  group_by(source, producer) %>% 
  filter(source != 'Total', year == 2020) %>% 
  ggplot() + 
  geom_col(aes(x = producer, y = generation/1000000, fill = source)) +
  labs(title = "US Energy Sources Among Producer Types in 2020", x= "Producer", y="Power Generation(TWH)")+
  theme(axis.text.x = element_text(angle = -15, vjust = 1, hjust = .1)) 

```


## Energy Sources amoung producer types TX
```{r}
# TX Energy Sources Among Producer Types in 2010
## This graph is meant to be used in comparison to the graph proceeding it 
## displaying the power source distributions among energy producers in Texas in 2010
energy %>% 
  group_by(source, producer, state) %>% 
  filter(source != 'Total', year == 2010, state == "TX") %>% 
  ggplot() + 
  geom_col(aes(x = producer, y = generation/1000000, fill = source)) +
  labs(title = "TX Energy Sources Among Producer Types in 2010", x= "Producer", y="Power Generation(TWH)")+
  theme(axis.text.x = element_text(angle = -15, vjust = 1, hjust = .1)) 
# TX Energy Sources Among Producer Types in 2020
## This graph is meant to be used in comparison to the graph proceeding it 
## displaying the power source distributions among energy producers in Texas in 2020
energy %>% 
  group_by(source, producer, state) %>% 
  filter(source != 'Total', year == 2020, state == "TX") %>% 
  ggplot() + 
  geom_col(aes(x = producer, y = generation/1000000, fill = source)) +
  labs(title = "TX Energy Sources Among Producer Types in 2020", x= "Producer", y="Power Generation(TWH)")+
  theme(axis.text.x = element_text(angle = -15, vjust = 1, hjust = .1)) 
```
