---
title: "United States Energy Generation 2001-2022"
author: "Kevin Sheard, John Beall, Andrew Vo"
date: "`r Sys.Date()`"
output: html_document
runtime: shiny
source: https://www.kaggle.com/datasets/kevinmorgado/us-energy-generation-2001-2022
---
# Introduction

Fossil fuels have been a huge detriment to our environment and we want to educate the audience with insight towards energy's evolution to combat fossil fuels and possibly influence one's decision to enact upon such information (with the exception of traffic blockers and vandals).In this data set we're looking at the 17 sustainable development goals of the United Nations. Within these 17 goals we saw that the UN's energy goal could be a potentially high value benefactor of data science to help curb the generation of fossil fuels and promote green energy. We chose this topic with the end goal of finding interesting discoveries that could provide actionable data to the every day consumer and help educate their decision to dictate their personal energy needs within current trends and available sources today. We hope that through our visualization of trending energy generation, production, and interactive comparisons an individual may be better informed to make their steps towards a greener future!


```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = F,
	message = FALSE,
	warning = FALSE
)
```


```{r}
# Load tidyverse for manipulating data
# Load ggplot2 for graphing data
# Load shiny and plotly for interactive graphs
# Load ggthemes for better aesthetics

library(tidyverse, warn.conflicts = F)
library(ggplot2, warn.conflicts = F)
library(plotly, warn.conflicts = F)
library(shiny, warn.conflicts = F)
library(ggthemes, warn.conflicts = F)
library(AER)
```

```{r}
#Read in the energy data
energy = read_csv("archive/organised_Gen.csv")

```

# Data Preparation
## Source
This data was collected from Kaggle titled [US Energy Generation 2001-2022](https://www.kaggle.com/datasets/kevinmorgado/us-energy-generation-2001-2022), which is a cleaned up data set adapted from the [US Energy Information Administration](https://www.eia.gov/electricity/). The Kaggle CSV was created by Kevin Morgado and is updated annually as new information is uploaded from the Energy Information Administration(EIA).

## Data Reading and Cleaning
This data was originally a CSV so we used the read_csv function to turn that CSV into a data set and stored it into an object called "energy". The original column names were messy to type so we created new column names. The energy data set also contained an erroneous ID column that was not necessary so we deleted it. Also, for the majority of our graphs we only used the "Total Electric Power Industry" producer for clarity. Lastly we created a new column called Date that was a combination of the month and year columns so that we could create time series graphs.

```{r}
#renaming the columns and removing the "ID" column
colnames(energy) = c("ID", "year", "month", "state", "producer", "source", "generation")
energy = energy %>% select(-`ID`)
#adjusting the dates so that they can used with a time series graph
energy$Date<-as.Date(with(energy,paste(year,month,Day = 1,sep="-")),"%Y-%m-%d")
```

## Glimpse of the Energy Dataset

```{r}
glimpse(energy)
```

| **Variable**   | **explanation**                                                                               |
|----------------|-----------------------------------------------------------------------------------------------|
| **year**       | the year that the power was generated                                                         |
| **month**      | the month that the power was generated                                                        |
| **state**      | the state that produced the power                                                             |
| **producer**   | the type of producer who created the power(industrial power, commercial power,...)            |
| **source**     | what created the power (nuclear, coal, petroleum,...)                                         |
| **generation** | the amount of power generated in Megawatt-hours                                               |
| **Date**       | the date in {year, month, day} format, the day is automatically set to the first of the month |


## US Power Generation


```{r}
#plot a graph for US avg. power generation per year from years 2001-may 2022
energy %>% 
  filter(producer == "Total Electric Power Industry", state == "US-TOTAL") %>% 
  ggplot() + 
  geom_point(aes(x = year, y = generation/1000000, color = source)) +
  #we will generation/1000000 to create Tera Watt Hours instead Mega Watt Hours for readability for the rest of the project
  geom_smooth(aes(x = year, y = generation/1000000, color = source)) +
  labs(title = "US average power generation per year", y = "Generation(TWH)", x = "Year") + theme_clean()
  
```

To get a better understanding of the sources of our power generation. 

## Accumulative US Power Generation over last 20+ Years

```{r fig.height=8, fig.width=5}
#setting figure boundaries to ensure readability
#plot a graph for accumulative US power generation per year from years 2001-may 2022 for each state
total_energy = energy %>% filter(producer == "Total Electric Power Industry", state != "US-TOTAL") 
fig = energy %>% 
  filter(producer == "Total Electric Power Industry", state != "US-TOTAL") %>% 
  group_by(year, month, state) %>% 
  summarize(total = sum(generation)/2) %>% 
  right_join(total_energy) %>%
  filter(source != "Total") %>% 
  arrange(total) %>% 
  ggplot() +
  geom_col(aes(x = state, y = generation/1000000, fill = source)) +
  #used columns and flipped coords to create clarity on graph
  labs(title = "US Power Generation by State", y = "Generation(TWH)", x = "State") +
  coord_flip() + theme_clean()

fig
```

To better understand the magnitude of power generated over the last 20+ years. We can notice TX is the largest producer of energy.

## Texas Power Generation

```{r}
#Display the Texas avg power generation per year from years 2001-may 2022 
energy %>% 
  filter(producer == "Total Electric Power Industry", state == "TX") %>% 
  ggplot() + 
  geom_point(aes(x = year, y = generation/1000000, color = source)) +
  geom_smooth(aes(x = year, y = generation/1000000, color = source)) +
  labs(title = "Texas average power generation per year", x = "Year", y = "Power Generation(TWH)") + theme_clean()

```

A look at how TX's power generation varies over the years, TX has a growing total power generation trend. 

```{r}
#Display the Texas avg power generation per month to show variation of the power usage over the course of a year
energy %>% 
  filter(producer == "Total Electric Power Industry", state == "TX") %>% 
  ggplot() + 
  geom_point(aes(x = month, y = generation/1000000, color = source)) +
  geom_smooth(aes(x = month, y = generation/1000000, color = source)) +
  labs(title = "Texas average power generation per month", x = "Month", y = "Power Generation(TWH)") + theme_clean()
```

A look at how TX's power generation varies over a year, but with all years shown and we notice that with the hotter months our energy use increases and decreases with cooler months.

## Texas Power Generation

```{r fig.height=20}
#choosing display height for better readability

ui <- fluidPage(
  # App title ----
  titlePanel("State(s) Total Power Generation Over Time"),
  # Sidebar layout with a input and output definitions ----
  sidebarLayout(
    # Sidebar panel for inputs ----
    sidebarPanel(
      position = "right",
      # checkbox for source
      checkboxGroupInput("sources", 
        label = h3("Source:"), 
        choices = unique(energy$source),
        selected = "Total"),
      # checkbox to determine facet status
      checkboxInput("facetStatus",
                    "Facet",
                    value = TRUE),
      # input selection for what state to graph on the top graph
      selectInput("state1",
                  label = "State(top)",
                  choices = unique(energy$state),
                  selected = "TX"),
      # input selection for what state to graph on the bottom graph
      selectInput("state2",
                  label = "State(bottom)",
                  choices = unique(energy$state),
                  selected = "CA")
    ),

    # Main panel for displaying outputs ----
    mainPanel(
      #output the top plot
      plotlyOutput("totalGenPlot",
                   width = "100%",
                   height = "100%"),
      #output the bottom plot
      plotlyOutput("totalGenPlot2")
    )
  )
)

server = function(input, output){
  #Use reactive so that we can interact with the shiny
  shinyEnergy = reactive({
    energy %>% filter(producer == "Total Electric Power Industry", state == input$state1, year != 2022, source == input$sources) %>% group_by(Date, source) %>% 
    summarize(total_generation = sum(generation)) %>% mutate(total_generation = total_generation / 1000000)
  })
  
  shinyEnergy2 = reactive({
    energy %>% filter(producer == "Total Electric Power Industry", state == input$state2, year != 2022, source == input$sources) %>% group_by(Date, source) %>% 
    summarize(total_generation = sum(generation)) %>% mutate(total_generation = total_generation / 1000000)
  })

  output$totalGenPlot = renderPlotly({
    #using plotly so Shiny behaves with our graph, and using fig2 as an object so that we can call up a way to use the Shiny
    fig2 = shinyEnergy() %>% 
    ggplot() +
    geom_line(aes(x= Date, y= total_generation, color = source)) +
    geom_smooth(aes(x = Date, y= total_generation, color = source)) +
    labs(x = "Date", y = "Power Generation(TWH)") 
    # facetEnergy is reactive so we can change its status, has source input and changes title based on the state check box
    facetEnergy = reactive({
      if(input$facetStatus == T){
        fig2 = fig2 + facet_wrap(~source)
      }
      else{
        fig2
      }
      fig2 = fig2 + labs(title = paste(input$state1, "Total Power Generation Over Time"))
    })
    
    fig2 = facetEnergy()
    #create the plotly
    ggplotly(fig2, tooltip = c("x", "y", "source"))
  })
  #second graph is constructed similarly to the graph above
  output$totalGenPlot2 = renderPlotly({
    fig2 = shinyEnergy2() %>% 
    ggplot() +
    geom_line(aes(x= Date, y= total_generation, color = source)) +
    geom_smooth(aes(x = Date, y= total_generation, color = source)) +
    labs(x = "Date", y = "Power Generation(TWH)") 
    
    facetEnergy = reactive({
      if(input$facetStatus == T){
        fig2 = fig2 + facet_wrap(~source)
      }
      else{
        fig2
      }
      fig2 = fig2 + labs(title = paste(input$state2, "Total Power Generation Over Time"))
    })
    
    fig2 = facetEnergy()
    
    ggplotly(fig2, tooltip = c("x", "y", "source"))
  })
}

shinyApp(ui = ui, server = server)
```

## Interesting Outliers

```{r}
#Noticed interesting outliers of data, and displayed it to show the magnitude of negative values of power generation.
energy %>% 
  filter(producer == "Total Electric Power Industry", state != "US-TOTAL", generation < -5000) %>% 
  ggplot() +
  geom_col(aes(x = state, y = generation/1000000, fill = source)) +
  labs(title = "States with negative power generation (<-5000)", x ="State", y="Power Generation(TWH)") + theme_clean()
```

We found data that doesn’t seem to fit, and after looking up these sources it is due to the need for input energy to be used for the generation of pumped storage. 


## US Power Generation

```{r}
#second shiny to display power generation for the US 
ui <- fluidPage(
  # App title ----
  titlePanel("US Total Power Generation Over Time"),
  # Sidebar layout with a input and output definitions ----
  sidebarLayout(
    # Sidebar panel for inputs ----
    sidebarPanel(
      position = "right",
      # checkbox for different sources
      checkboxGroupInput("sources", 
        label = h3("Source:"), 
        choices = unique(energy$source),
        selected = "Total"),
      # checkbox for whether to facet graph or not
      checkboxInput("facetStatus",
                    "Facet",
                    value = TRUE),
  
      hr(),
      fluidRow(column(3, verbatimTextOutput("value")))
    ),

    # Main panel for displaying outputs ----
    mainPanel(
      # Output: Verbatim text for data summary ----
      verbatimTextOutput("summary"),
      plotlyOutput("totalGenPlot")
    )
  )
)

server = function(input, output){
  # reactive in order to get what source to filter by
  shinyEnergy = reactive({
    energy %>% filter(producer == "Total Electric Power Industry", state != "US-TOTAL", year != 2022, source == input$sources) %>% group_by(Date, source) %>% 
    summarize(total_generation = sum(generation)) %>% mutate(total_generation = total_generation / 1000000)
  })
  
  
  #render the plot
  output$totalGenPlot = renderPlotly({
    fig2 = shinyEnergy() %>% 
    ggplot() +
    geom_line(aes(x= Date, y= total_generation, color = source)) +
    geom_smooth(aes(x = Date, y= total_generation, color = source)) +
    labs(title = "Total Power Generation Over Time", x = "Date", y = "Power Generation(TWH)") 
    
    # reactive function to get facet status
    facetEnergy = reactive({
      if(input$facetStatus == T){
        fig2 + facet_wrap(~source)
      }
      else{
        fig2
      }
    })
    
    fig2 = facetEnergy()
    # create a plotly from ggplot
    ggplotly(fig2, tooltip = c("x", "y", "source"))
  })
}

shinyApp(ui = ui, server = server)
```

## Producer Data Analysis

```{r}
#tables to display producer columns and energy sources
energy %>% count(producer)

energy %>% count(source)
```

A look into how we organized the data for the following producer data. 

## Texas/US Power Distribution Producer Type

```{r}
# US Power Source Distribution Per Producer Type. 
## This is meant to show the level of usage among energy source types used within US Energy industry categories. 
energy %>%
  filter(producer != "Total Electric Power Industry", source!="Total") %>% 
  ggplot() + 
  geom_col(aes(x = producer, y = generation/1000000, fill = source)) +
  labs(title = "US Power Distribution", x= "Producer", y= "Power Generation(TWH)") +
  theme(axis.text.x = element_text(angle = -25, vjust = .5, hjust = .1)) 
#theme adjusted to create readable labels
```

To better understand where the distribution of power is coming from for the US in general. This graph is a summation of all energy source usage despite increases or decreases from the changes over time. 
 

```{r}
# TX Power Source Distribution Per Producer Type.
## This is meant to display the same comparison as the US graph, but in Texas. 
energy %>% 
  filter(producer != "Total Electric Power Industry", state == 'TX', source != 'Total') %>%
  ggplot() + 
  geom_col(aes(x = producer, y = generation/1000000, fill = source)) +
  labs(title = "Texas Power Distribution", x= "Producer", y= "Power Generation(TWH)") +
  theme(axis.text.x = element_text(angle = -25, vjust = .5, hjust = .1)) 
#theme adjusted to create readable labels
```

To better understand where the distribution of power is coming from for TX specifically, we can notice that TX more heavily relies on Independent Power Producers. Probably due to our independent energy grid. This graph uses the same standards as the US graph above without regard of increases or decreases over the time frame of this data set.

## Source Power Generation Per Year

```{r}
# US Power Source Generation Per Year
energy %>% 
  filter(producer != "Total Electric Power Industry", state == "US-TOTAL", source != 'Total', year <= 2021) %>% 
  ggplot() + 
  geom_col(aes(x = year, y = generation/1000000, fill = source)) +
  labs(title = "US Power Generation Per Year", x = "Year", y = "Power Generation(TWH)")

```

This is meant to show the change of power sources used in the US on a chronological scale, here we notice the decline in coal, and growth in natural gas, wind, and solar.

```{r}
# TX Power Source Generation Per Year
energy %>% 
  filter(producer != "Total Electric Power Industry", state == "TX", source != 'Total', year <= 2021) %>% 
  ggplot() + 
  geom_col(aes(x = year, y = generation/1000000, fill = source)) +
  labs(title = "TX Power Generation Per Year", x = "Year", y = "Power Generation(TWH)")
```

This is meant to show the change of power sources used on a chronological scale in Texas, similarly to US but wind has massive growth comparatively.


## Total Power Distribution Over Months

```{r}
# Total US power distribution over months
energy %>% 
  filter(producer == "Total Electric Power Industry", state == 'US-TOTAL', source == 'Total') %>% 
  ggplot(aes(group = month, x = month, y = generation/1000000, fill = month))+
  geom_boxplot() +
  scale_x_continuous(breaks = c(1:12)) + labs(title= "Total US Power distribution over months", x= "Months", y="Power Generation(TWH)" ) 
```

Graph is meant to create a visual on power generation over months and used as a control for comparison for states and their power generation over months. This is to set precedent for the next pair of graphs.
 
 
```{r}
# TX power distribution over months
energy %>% 
  filter(producer == "Total Electric Power Industry", state == 'TX', source == 'Total') %>% 
  ggplot(aes(group = month, x = month, y = generation/1000000, fill = month))+
  geom_boxplot() +
  scale_x_continuous(breaks = c(1:12))+
  labs(title = 'TX Power Distribution Over Months', x= "Months", y="Power Generation(TWH)") 
```

Graph is meant to show Texas power usage in comparison to US. This is to set precedent for the next pair of graphs.


## Power Distributions over months for 2021

```{r}
# US Power Distributions over months for 2021
energy %>% 
  filter(producer != "Total Electric Power Industry", state == 'US-TOTAL', source != 'Total', year == 2021) %>% 
  ggplot()+
  geom_col(aes(x = month, y = generation/1000000, fill = source)) +
  labs(title = 'US Power Distribution Per Month', x= "Months", y="Power Generation(TWH)")+
  scale_x_continuous(breaks = c(1:12))
```

This graph shows the use of power sources in the US over a twelve month period in the year 2021, notice the increase of solar between May and October.


```{r}
# TX Power Distributions over months for 2021
energy %>% 
  filter(producer != "Total Electric Power Industry", state == 'TX', source != 'Total', year == 2021) %>% 
  ggplot()+
  geom_col(aes(x = month, y = generation/1000000, fill = source)) +
  labs(title = 'Texas Power Distribution Per Month', x= "Months", y="Power Generation(TWH)")+
  scale_x_continuous(breaks = c(1:12))
```

This graph shows the use of power sources in the TX over a twelve month period in the year 2021, similarly to the US in general but once again a greater percentage of wind power utilized. 


## Energy Sources among producer types US

```{r}
# US Energy Sources Among Producer Types in 2010
## displaying the power source distributions among energy producers in the US in 2010
energy %>% 
  group_by(source, producer) %>% 
  filter(source != 'Total', year == 2010) %>% 
  ggplot() + 
  geom_col(aes(x = producer, y = generation/1000000, fill = source)) +
  labs(title = "US Energy Sources Among Producer Types in 2010", x= "Producer", y="Power Generation(TWH)")+
  theme(axis.text.x = element_text(angle = -15, vjust = 1, hjust = .1)) 
#theme adjusted to create readable labels
```

This graph is looking at how the distribution of energy was sourced between producers in 2010 for the US to create a baseline for future change. 


```{r}
# US Energy Sources Among Producer Types in 2020
## displaying the power source distributions among energy producers in the US in 2020
energy %>% 
  group_by(source, producer) %>% 
  filter(source != 'Total', year == 2020) %>% 
  ggplot() + 
  geom_col(aes(x = producer, y = generation/1000000, fill = source)) +
  labs(title = "US Energy Sources Among Producer Types in 2020", x= "Producer", y="Power Generation(TWH)")+
  theme(axis.text.x = element_text(angle = -15, vjust = 1, hjust = .1)) 
#theme adjusted to create readable labels
```

This graph is comparing the distribution of energy sourced from 2020 in the US to acknowledge change, such as the decrease in coal, increase in natural gas, wind, and solar. 


## Energy Sources amoung producer types TX

```{r}
# TX Energy Sources Among Producer Types in 2010
## displaying the power source distributions among energy producers in Texas in 2010
energy %>% 
  group_by(source, producer, state) %>% 
  filter(source != 'Total', year == 2010, state == "TX") %>% 
  ggplot() + 
  geom_col(aes(x = producer, y = generation/1000000, fill = source)) +
  labs(title = "TX Energy Sources Among Producer Types in 2010", x= "Producer", y="Power Generation(TWH)")+
  theme(axis.text.x = element_text(angle = -15, vjust = 1, hjust = .1)) 
#theme adjusted to create readable labels
```

This graph is looking at how the distribution of energy was sourced between producers in 2010 for Texas to create a baseline for future change


```{r}
# TX Energy Sources Among Producer Types in 2020
## displaying the power source distributions among energy producers in Texas in 2020
energy %>% 
  group_by(source, producer, state) %>% 
  filter(source != 'Total', year == 2020, state == "TX") %>% 
  ggplot() + 
  geom_col(aes(x = producer, y = generation/1000000, fill = source)) +
  labs(title = "TX Energy Sources Among Producer Types in 2020", x= "Producer", y="Power Generation(TWH)")+
  theme(axis.text.x = element_text(angle = -15, vjust = 1, hjust = .1)) 
#theme adjusted to create readable labels
```

This graph is comparing the distribution of energy sourced from 2010 in Texas to acknowledge change such as the decrease in coal, increase in natural gas, wind, and solar.

# Summarization
#### Summarize the problem statement
Investigate the sources of power generation in the US to find relatable trends to the climate action and clean energy aspects of the United Nations Sustainable Developmental Goals, and if this data could be used to work towards the SDG's.

#### Addressing the problem statement
By analyzing the different sources of power generation, the magnitudes of generation, and the distribution of the power across needs and relating them with those same factors over time in search of trends.


#### Interesting insights
Pumped storage has caveats due to the nature of using energy to store energy to a net loss, but useful when energy generation supply supersedes demand. Solar power is touted as a huge alternative energy source and it has significantly grown but wind power has outpaced its power generation. Coal, has drastically reduced in use, while natural gas has essentially replaced it. Nuclear, conventional hydroelectric, and geothermal have all stayed relatively the same over the past 20 years. 


#### Implications 
We ultimately see in this analysis the major decline of coal, the budding rise of green energy in wind and solar, and the titanic gain in natural gas and while natural gas is a much cleaner burning source compared to the coal it replaces it is a fossil fuel nonetheless. The US is headed towards a better direction but the proportion of energy sourced from fossil fuels is still much higher than desired if we are to combat climate change. We are making noticeable steps for a more sustainable future.

#### Limitations
For general purposes this dataset is fit to serve, but there is much more room for discovery within. For instance there is more room for analysis on such as deeper comparisons between energy sources and US comparisons to states. Another possibility is the ability for one to dive deeper and hyper-focus on a states particular energy needs. Perhaps an analysis of industries and their differentiation would be highly informative as well. While we were not overly limited in our findings, a great way to improve upon them would be to gather relatable economic data on the energy sourced, and thus how much was spent and how efficient that money spent was. Another thing that could be done to improve upon this is create an approximation of the pollution and/or GHG(greenhouse gases) created during the use of these. 
