---
title: "Final Project"
author: John Beall
date: November 27, 2022
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# Load tidyverse for manipulating data
#Load ggplot2, and plotly for graphing data
library(tidyverse, warn.conflicts = F)
library(ggplot2, warn.conflicts = F)
library(plotly, warn.conflicts = F)
```

```{r}
#Read in the energy data
energy = read_csv("archive/organised_Gen.csv")
glimpse(energy)

```


```{r}
#renaming the columns and removing the "ID" column
colnames(energy) = c("ID", "year", "month", "state", "producer", "source", "generation")
energy = energy %>% select(-`ID`)
```

```{r}
energy$Date<-as.Date(with(energy,paste(year,month,Day = 1,sep="-")),"%Y-%m-%d")
energy
```

##US total power generation per year
```{r}
energy %>% 
  filter(producer == "Total Electric Power Industry", state == "US-TOTAL") %>% 
  ggplot() + 
  geom_point(aes(x = year, y = generation, color = source)) +
  geom_smooth(aes(x = year, y = generation, color = source)) +
  labs(title = "US total power generation per year")
  
```

## Total power generation per state
```{r}
energy %>% 
  filter(producer == "Total Electric Power Industry", state != "US-TOTAL") %>% 
  ggplot() +
  geom_point(aes(x = state, y = generation, color = source)) +
  labs(title = "Total power generation per state")
```

## Total Power Generation Per Year
```{r}
energy %>% filter(producer == "Total Electric Power Industry", state != "US-TOTAL", source != "Total", year != 2022) %>% group_by(source, year) %>% 
  summarize(total_generation = sum(generation)) %>% mutate(total_generation = total_generation / 1000000) %>% 
  ggplot() +
  geom_col(aes(x= year, y= total_generation)) +
  geom_smooth(aes(x = year, y= total_generation)) +
  facet_wrap(~source) +
  labs(title = "Total Power Generation Per Year", x = "Year", y = "Power Generation(TWH)")
```

## Total Solar Power Generation Per Year

```{r}
energy %>% filter(producer == "Total Electric Power Industry", state != "US-TOTAL", source == "Solar Thermal and Photovoltaic", year != 2022) %>% group_by(year) %>% 
  summarize(total_generation = sum(generation)) %>% mutate(total_generation = total_generation / 1000000) %>% 
  ggplot() +
  geom_col(aes(x= year, y= total_generation)) +
  geom_smooth(aes(x = year, y= total_generation)) +
  labs(title = "Total Solar Power Generation Per Year", x = "Year", y = "Power Generation(TWH)")
```

## Total Wind Power Generation Per Year

```{r}
energy %>% filter(producer == "Total Electric Power Industry", state != "US-TOTAL", source == "Wind", year != 2022) %>% group_by(year) %>% 
  summarize(total_generation = sum(generation)) %>% mutate(total_generation = total_generation / 1000000) %>% 
  ggplot() +
  geom_col(aes(x= year, y= total_generation)) +
  geom_smooth(aes(x = year, y= total_generation)) +
  labs(title = "Total Wind Power Generation Per Year", x = "Year", y = "Power Generation(TWH)")
```

## Total Petroleum Power Generation Per Year

```{r}
energy %>% filter(producer == "Total Electric Power Industry", state != "US-TOTAL", source == "Petroleum", year != 2022) %>% group_by(year) %>% 
  summarize(total_generation = sum(generation)) %>% mutate(total_generation = total_generation / 1000000) %>% 
  ggplot() +
  geom_col(aes(x= year, y= total_generation)) +
  geom_smooth(aes(x = year, y= total_generation)) +
  labs(title = "Total Petroleum Power Generation Per Year", x = "Year", y = "Power Generation(TWH)")
```

## Total Power Generation Over Time

```{r}
energy %>% filter(producer == "Total Electric Power Industry", state != "US-TOTAL", year != 2022, source == c("Solar Thermal and Photovoltaic", "Petroleum", "Wind")) %>% group_by(Date, source) %>% 
  summarize(total_generation = sum(generation)) %>% mutate(total_generation = total_generation / 1000000) %>%
  ggplot() +
  geom_line(aes(x= Date, y= total_generation, color = source)) +
  geom_smooth(aes(x = Date, y= total_generation, color = source)) +
  labs(title = "Total Power Generation Over Time", x = "Date", y = "Power Generation(TWH)")
```

## Accumulative Power Generation Over Time (2001-2022)

```{r}
energy %>% filter(producer == "Total Electric Power Industry", state != "US-TOTAL", source == "Solar Thermal and Photovoltaic") %>%
  ggplot() +
  geom_col(aes(x= state, y= generation)) + labs(title = "Accumulative Power Generation Over Time (2001-2022)", x = "States", y = "Generation (MWH)") +
  coord_flip() 
```

## U.S. Total Power Generated
```{r}
energy %>% 
  ggplot() +
  geom_col(aes(x = year, y = generation, fill = source)) +
  facet_wrap(vars(source)) + labs(title= "U.S. Total Power Generated", x = "Year(2001-2022)", y= "Generation(MWH)")
  
```

## States with negative power generation (<-5000)

```{r}
energy %>% 
  filter(producer == "Total Electric Power Industry", state != "US-TOTAL", generation < -5000) %>% 
  ggplot() +
  geom_point(aes(x = state, y = generation, color = source)) +
  labs(title = "States with negative power generation (<-5000)", x = "State", y= "Generation(MWH)")
```


US goal is to install an average of 30 GW of solar capacity per
year between now and 2025 and 60 GW per year from 2025-2030.
https://www.renewable-ei.org/pdfdownload/activities/01_Key_AlejandroMoreno.pdf


```{r, include =FALSE}
total_energy = energy %>% filter(producer == "Total Electric Power Industry", state != "US-TOTAL") 
fig = energy %>% 
  filter(producer == "Total Electric Power Industry", state != "US-TOTAL") %>% 
  group_by(year, month, state) %>% 
  summarize(total = sum(generation)/2) %>% 
  right_join(total_energy) %>%
  filter(source != "Total") %>% 
  ggplot() +
  geom_col(aes(x = state, y = generation, fill = source))

ggplotly(fig, tooltip = "y")
```


## Texas total power generation per year
```{r}
energy %>% 
  filter(producer == "Total Electric Power Industry", state == "TX") %>% 
  ggplot() + 
  geom_point(aes(x = year, y = generation, color = source)) +
  geom_smooth(aes(x = year, y = generation, color = source)) +
  labs(title = "Texas total power generation per year", x = "Year(2001-2022)", y= "Generation(MWH)")
```


## Texas total power generation per month
```{r}
energy %>% 
  filter(producer == "Total Electric Power Industry", state == "TX") %>% 
  ggplot() + 
  geom_point(aes(x = month, y = generation, color = source)) +
  geom_smooth(aes(x = month, y = generation, color = source)) +
  labs(title = "Texas total power generation per month", x = "Month", y= "Generation(MWH)")
```


## Texas Total Wind Power Generation Per Year
```{r}
energy %>% filter(producer == "Total Electric Power Industry", state == "TX", source == "Wind", year != 2022) %>% group_by(year) %>% 
  summarize(total_generation = sum(generation)) %>% mutate(total_generation = total_generation / 1000000) %>% 
  ggplot() +
  geom_col(aes(x= year, y= total_generation)) +
  geom_smooth(aes(x = year, y= total_generation)) +
  labs(title = "Texas Total Wind Power Generation Per Year", x = "Year", y = "Power Generation(TWH)")
```


## Texas Total Petroleum Generation Per Year
```{r}
energy %>% filter(producer == "Total Electric Power Industry", state == "TX", source == "Petroleum", year != 2022) %>% group_by(year) %>% 
  summarize(total_generation = sum(generation)) %>% mutate(total_generation = total_generation / 1000000) %>% 
  ggplot() +
  geom_col(aes(x= year, y= total_generation)) +
  geom_smooth(aes(x = year, y= total_generation)) +
  labs(title = "Texas Total Petroleum Generation Per Year", x = "Year", y = "Power Generation(TWH)")
```


## Texas Total Solar Power Generation Per Year
```{r}
energy %>% filter(producer == "Total Electric Power Industry", state == "TX", source == "Solar Thermal and Photovoltaic", year != 2022) %>% group_by(year) %>% 
  summarize(total_generation = sum(generation)) %>% mutate(total_generation = total_generation / 1000000) %>% 
  ggplot() +
  geom_col(aes(x= year, y= total_generation)) +
  geom_smooth(aes(x = year, y= total_generation)) +
  labs(title = "Texas Total Solar Power Generation Per Year", x = "Year", y = "Power Generation(TWH)")
```


## Total Power Generation Over Time
```{r}
energy %>% filter(producer == "Total Electric Power Industry", state == "TX", year != 2022, source == c("Solar Thermal and Photovoltaic", "Petroleum", "Wind")) %>% group_by(Date, source) %>% 
  summarize(total_generation = sum(generation)) %>% mutate(total_generation = total_generation / 1000000) %>%
  ggplot() +
  geom_line(aes(x= Date, y= total_generation, color = source)) +
  geom_smooth(aes(x = Date, y= total_generation, color = source)) +
  labs(title = "Total Power Generation Over Time", x = "Date", y = "Power Generation(TWH)")
```